---
title: "2016 US Presidential Primary Season"
author: "Salil Gupta"
date: "April 5, 2016"
output: html_document
---

This was my input for the following Kaggle Competition: <https://www.kaggle.com/benhamner/2016-us-election>

The dataset was primary results from the "early" states for the 2016 US Presidential Election.
Also, included with the data were "County Facts", variables that described the population in that County.  

#Libraries Used

Library | Description
------------- | -------------
Chropleth | A library for plotting US County and State data
RPart | A library usued for its recursive partioning algorithms
GGPlot2 | Grammar of Graphics Plotting library
Reshape2 | Pivoting data from long to wide
caret | ML Library
RSQLite | Connect to SQLite tables
DPlyr | The Data Wrangling Package
SQLDF | Query dfs like tables
RColorBrewer | Produce Better Looking Trees






```{r eval=TRUE, message = FALSE , echo = FALSE, warning = FALSE}

library(maps)
library(corrplot)
library(ggplot2)
library(dplyr)
library(reshape2)
library(RSQLite)
library(choroplethr)
library(choroplethrMaps)
library(rpart)
library(rpart.plot)
library(datasets)
library(caret)
library(sqldf)
library(RColorBrewer)
library(rattle)
```

## Connect to SQLite dB 

```{r eval=TRUE, message = FALSE , echo = TRUE, warning = FALSE}
dbpath = "H:/PE/2016_presidential_election/database.sqlite"
db <- dbConnect(dbDriver("SQLite"), dbpath)
dbListTables(db)

countyfacts <- dbGetQuery(db, "select *  from county_facts")
cfd <- dbGetQuery(db, "select *  from county_facts_dictionary")
presults <- dbGetQuery(db, "select *  from primary_results")
rsPR <- data.frame(presults$fips,presults$party,presults$candidate, presults$fraction_votes)
```

## Pivot the Candidate's Percentage of Votes to Columns for each County Code 
```{r eval=TRUE}
colnames(rsPR) <- c("fips","party","cand","perVotes")
newdf <- dcast(rsPR,fips ~ cand , value.var = "perVotes")

mydf <- merge(newdf,countyfacts, by = "fips")
mydf$ID <- rownames(mydf)
##Derived column 
##head(mydf,1)
```

## Derived Columns to get County Winners, and Binary Fields for Candidate of Interest (Trump)
```{r eval=TRUE}
demdf <- data.frame(mydf$`Hillary Clinton`, mydf$`Bernie Sanders`)
demwinners <- colnames(demdf)[apply(demdf,1,which.max)]

repubdf <- data.frame(mydf$`Ben Carson`, mydf$`Carly Fiorina`, mydf$`Chris Christie`, mydf$`Jeb Bush`, mydf$`John Kasich`, mydf$`Marco Rubio`, mydf$`Ted Cruz`, mydf$`Donald Trump`, mydf$`Ted Cruz`, mydf$`Mike Huckabee` ,mydf$`Rand Paul`, mydf$`Rick Santorum`)
rwinners <- colnames(repubdf)[apply(repubdf,1,which.max)]

dwdf <- data.frame(demwinners)
dwdf$ID <- rownames(dwdf)

rwdf <- data.frame(rwinners)
rwdf$ID <- rownames(rwdf)

winners <- merge(rwdf,dwdf, by = "ID")
mydf <- merge(winners,mydf, by = "ID")
BinaryCandidateWin <- sqldf("select ID, 
                   
                   case when rwinners = 'mydf..Donald.Trump.' then 1 else 0 end as TrumpWin,
                   case when demwinners = 'mydf..Hillary.Clinton.' then 1 else 0 end as HilWin

                   from mydf ")
mydf <- merge(BinaryCandidateWin,mydf, by = "ID")

```




## Map Plots :  Income and Trump Percentage 
```{r eval=TRUE, message = FALSE , echo = TRUE, warning = FALSE}

 
  df <- data.frame(mydf$fips, mydf$INC110213 )
  colnames(df) <- cbind("region", "value")
  
  
  county_choropleth(df)
  
  
  df <- data.frame(mydf$fips, mydf$`Donald Trump` )
  colnames(df) <- cbind("region", "value")
  
  
  county_choropleth(df)
  
 ## GG Plot

  ggplot(aes(x=RHI725214,y=AGE775214,color=mydf$rwinners),data=mydf)+geom_jitter()
  dropcols <- c("rwinners", "demwinners","area_name","state_abbreviation" )
cordf <- mydf[ , !names(mydf) %in% dropcols ]

i <- sapply(cordf, is.character)
cordf[i] <- lapply(cordf[i], as.numeric)


mycors <- cor(cordf, use="complete.obs", method="kendall") 

##myelectioncorplot <- corrplot(mycors,method = "circle")

treeFormu <- formula(TrumpWin ~  PST045214 + 
                     PST040210 +                      PST120214 +                      POP010210 +                      AGE135214 + 
                     AGE295214 +                      AGE775214 +                      SEX255214 +                      RHI125214 + 
                     RHI225214 +                      RHI325214 +                      RHI425214 +                      RHI525214 + 
                     RHI625214 +                      RHI725214 +                      RHI825214 +                      POP715213 + 
                     POP645213 +                      POP815213 +                      EDU635213 +                      EDU685213 + 
                     VET605213 +                      LFE305213 +                      HSG010214 +                      HSG445213 + 
                     HSG096213 +                      HSG495213 +                      HSD410213 +                      HSD310213 + 
                     INC910213 +                      INC110213 +                      PVY020213 +                      BZA010213 + 
                     BZA110213 +                      BZA115213 +                      NES010213 +                      SBO001207 + 
                     SBO315207 +                      SBO115207 +                      SBO215207 +                      SBO515207 + 
                     SBO415207 +                      SBO015207 +                      MAN450207 +                      WTN220207 + 
                     RTN130207 +                      RTN131207 +                      AFN120207 +                      BPS030214 + 
                     LND110210 +                      POP060210 )
 set.seed(1001)
rprimindex <- createDataPartition(y=mydf$rwinners, times = 1, p =.5, list = F)
mydf.train <-mydf[rprimindex, ]
mydf.test  <-mydf[-rprimindex, ]
set.seed(1)
treeModel <- rpart(treeFormu,  data=mydf.train)


trumpPred.train<- factor(format(round(predict(treeModel, mydf.train))))
trumpPred.test<- factor(format(round(predict(treeModel, mydf.test))))
##table(trumpPred, mydf.test$HilWin)

confusionMatrix(trumpPred.train, mydf.train$TrumpWin)
confusionMatrix(trumpPred.test, mydf.test$TrumpWin)


fancyRpartPlot(treeModel)
  
```