---
title: 'Kaggle Script:  American Community Survey'
author: "Salil Gupta"
date: "April 7, 2016"
output: html_document
---

This was my input for the following Kaggle Competition: <https://www.kaggle.com/benhamner/2016-us-election>

The data consisted of two different data sets, one on the grain of individual people, the other on indvidiual households.  
Also, included with the data were "County Facts", variables that described the population in that County.  

#Libraries Used

Library | Description
------------- | -------------
Data.Table | A library for plotting US County and State data
GGPlot2 | Grammar of Graphics Plotting library


```{r eval=TRUE, message = FALSE , echo = FALSE, warning = FALSE}

library(data.table)

library(ggplot2)
library(sqldf)
```

##ST STate




Variable | Description
------------- | -------------
ST | State
PWGTP | A library for plotting US County and State data
AGEP | AGE
CIT | CITIZENSHIP
CITWP | A library for plotting US County and State data
COW | Class of Worker
DEAR | A library for plotting US County and State data
DEYE | A library for plotting US County and State data
ENG | English Skills
MAR | Marital Status
SCH | Current Student Status
SCHG | A library for plotting US County and State data
SCHL | A library for plotting US County and State data
SEX | Gender
WAGP | A library for plotting US County and State data
ANC1P | A library for plotting US County and State data
POBP | A library for plotting US County and State data
VPS | Veteran Status
ESR | A library for plotting US County and State data
PINCP | Income
RAC1P | Race


| Grammar of Graphics Plotting library



## Data.Table package is used to only load the columns we need from the external dataset
```{r eval=TRUE, message = FALSE , echo = FALSE, warning = FALSE}
popCols <- c("ST", "PWGTP", "AGEP" ,"CIT", "CITWP", "COW", "DEAR", "DEYE", "ENG", "MAR" ,  "SCH", "SCHG", "SCHL", "SEX", "WAGP", "ANC1P", "POBP" ,"VPS", "ESR", "PINCP" ,"RAC1P")

popDF <- fread("H:\\Py\\csv_pus\\ss13pusa.csv", select  = popCols)


popDF$ST <- factor(popDF$ST)
popDF$CIT <- factor(popDF$CIT)
popDF$COW <- factor(popDF$COW)
popDF$DEAR <- factor(popDF$DEAR)
popDF$DEYE <- factor(popDF$DEYE)
popDF$ENG <- factor(popDF$ENG)
popDF$MAR <- factor(popDF$MAR)
popDF$SCH <- factor(popDF$SCH)
popDF$SCHGFactor <- factor(popDF$SCHG)
popDF$SCHLFactor <- factor(popDF$SCHL)
popDF$SEX <- factor(popDF$SEX)
popDF$ESR <- factor(popDF$ESR)
popDF$POBP <- factor(popDF$POBP)
popDF$VPS <- factor(popDF$VPS)
popDF$RAC1P <- factor(popDF$RAC1P)

levels(popDF$CIT) <- c("Born in US", "Born in US Terri", "American Born abroad", "Naturalized", "Non-Citizen")
levels(popDF$COW) <- c("For-Profit" , "Non-Proft", "Local Gov", "State Gov", "Fed Gov", "Self-Employed A", "Self-Employed B", "Without Income - Family Biz or Farm", "Unemployed")
levels(popDF$ENG) <- c("Very Well", "Well", "Not Well", "None")
levels(popDF$MAR) <- c("Married", "Widowed", "Divorced", "Seperated", "Never Married or under 15")
levels(popDF$SCH) <- c("No", "Yes Public", "Yes Private")
levels(popDF$SEX) <- c("Male", "Female")
levels(popDF$ESR) <- c("Civilian, at work", "Civilian, Not at Work", "Unemployed", "Armed Forces, at work", "Armed Forces, Not at work", "Non-Labor Force")
levels(popDF$ST) <-  c("alabama", "alaska", "arizona", "arkansas", "california", "colorado", "connecticut",
                         "delaware", "district of columbia", "florida", "georgia", "hawaii", "idaho", "illinois",
                         "indiana", "iowa", "kansas", "kentucky", "louisiana", "maine", "maryland", "massachusetts",
                         "michigan", "minnesota", "mississippi", "missouri", "montana", "nebraska", "nevada",
                         "new hampshire", "new jersey", "new mexico", "new york", "north carolina", "north dakota",
                         "ohio", "oklahoma", "oregon", "pennsylvania", "rhode island",
                         "south carolina", "south dakota", "tennessee", "texas", "utah", "vermont", "virginia",
                         "washington", "west virginia", "wisconsin", "wyoming", "puerto rico")

levels(popDF$RAC1P) <- c("White", "Black", "American Indian", "Alaska Native",  "American Indian - Tribe Specific", "Asian", "Pacific Islander", "Other", "Two or More Races")
```

## Case Statement to Recode School Grade Attainment into Degree Buckets using SQLDF package (Very effective for derived columns, if this SSIS developer  says so)
\
```{r eval=TRUE, message = FALSE , echo = TRUE, warning = FALSE}
popDF$SCHLRecode <- sqldf(
  "
  select case 
  when SCHL <= 15 then 'NonGrad' 
  when SCHL = 16 then 'HS' 
  when SCHL = 17 then 'GED' 
  when SCHL <= 19 and SCHL >= 18 then 'College Drop Out' 
  when SCHL = 20 then 'Associate'
  when SCHL = 21 then 'Bachelor'
  when SCHL = 22 then 'Master'
  when SCHL = 23 then 'Professional'
  when SCHL = 24 then 'Doctorate'
  
  else 'NULL' end 

  as SCHLRecode

  from popDF"
  
)
```
```{r eval=TRUE, message = FALSE , echo = TRUE, warning = FALSE}
popDF$SCHLRecode <- sqldf(
  "
  select case 
  when SCHL <= 15 then 'NonGrad' 
  when SCHL = 16 then 'HS' 
  when SCHL = 17 then 'GED' 
  when SCHL <= 19 and SCHL >= 18 then 'College Drop Out' 
  when SCHL = 20 then 'Associate'
  when SCHL = 21 then 'Bachelor'
  when SCHL = 22 then 'Master'
  when SCHL = 23 then 'Professional'
  when SCHL = 24 then 'Doctorate'
  
  else 'NULL' end 

  as SCHLRecode

  from popDF"
  
)
```